Towards Verifying the Bitcoin-S Library

### 2019-06-26 paper/paper.tex
work on section 2

%\usepackage{newtxtext}
%\usepackage{newtxmath}
\usepackage[zerostyle=b,scaled=.9]{newtxtt}

@@

\paragraph{Outline.} In the next section we try to verify the property
that a regular (non-coinbase) transaction cannot generate new
coins. We call it the \emph{no-inflation property}. Trying to verify
it, we uncover and fix a bug in the bitcoin-s library. We find that
there is too much code involved that lies outside of the supported
fragment to currently make this verification feasible. Instead, in
Section~3 we turn to a simpler property to verify. The simplest
possible property we could think of was the fact that adding zero
satoshis to a given amount of satoshis yields the given amount of
satoshis. We call it the \emph{addition-with-zero property}. Here as
well we see that a significant part of the code lies outside of the
supported fragment. However, after a series of equivalent
transformations we arrive at code that we successfully
verify. \todo{conclusions: what now?}

@@

\paragraph{Creating a Transaction} Some code in this section is copied or adapted from the Bitcoin-S-Core
transaction builder example \cite{BitcoinSCore:txbuilderexample}.
Bitcoin-S-Core has a bitcoin transaction builder class with the
following constructor:
\begin{lstlisting}[style=scala]
  BitcoinTxBuilder(
    destinations: Seq[TransactionOutput], 
    utxos: BitcoinTxBuilder.UTXOMap,      
    feeRate: FeeUnit,                     
    changeSPK: ScriptPubKey,              
    network: BitcoinNetwork               
  ): Future[BitcoinTxBuilder]
\end{lstlisting}

The return type Future does not make sense here, since the
implementation calls either Future.successful or Future.fromTry which
returns an already resolved Future.  This might be for future
purposes.

Now we create a transaction.

First, we need some money.  Thus, we create a fake transaction with
one single output.  This transaction can be parsed from the bitcoin
network, but we create one manually in order to see this process.
\begin{lstlisting}[style=scala]
  val privKey = ECPrivateKey.freshPrivateKey
  val creditingSPK = P2PKHScriptPubKey(pubKey = privKey.publicKey)

@@

On line one and two we create a new keypair to sign the next
transaction and have a scriptPubKey where the bitcoins are.  This is
our keypair.  So the money is transferred to our public key.  Line
four specifies the amount of satoshis we have in the transaction.
Then we create the actual transaction from line 6 to 13.

Now that we have some bitcoins, we create the new transaction where we
want to spend them.

First, we need some out points.  They point to outputs of previous
transactions.  We use the index zero, because the previous transaction
has only one output that becomes the first index zero.  If there were
two previous outputs, the second output would become the index 1 and
so on.
\begin{lstlisting}[style=scala]
  val outPoint = TransactionOutPoint(prevTx.txId, UInt32.zero)

@@

  val utxos = List(utxoSpendingInfo)
\end{lstlisting}

This utxos are the inputs of our transaction.

Second, we need destinations to spend the bitcoins to.
For the sake of convenience we create only one.

\begin{lstlisting}[style=scala]
  val destinationAmount = Satoshis(Int64(5000))

@@

  val destinations = List(
    TransactionOutput(currencyUnit = destinationAmount, scriptPubKey = destinationSPK)
  )
\end{lstlisting}

We spend 5000 satoshis to the newly created random public key.

Finally, we define the fee rate in satoshis per one byte transaction
size as well as some bitcoin network parameters.  The bitcoin network
parameters are not important, so we use some static values normally
used when testing.

\begin{lstlisting}[style=scala]
  val feeRate = SatoshisPerByte(Satoshis.one)

  val networkParams = RegTest // some static values for testing
\end{lstlisting}

Now lets build the transaction with those data.
\begin{lstlisting}[style=scala]
  val txBuilderF: Future[BitcoinTxBuilder] = BitcoinTxBuilder(
    destinations = destinations, // where to send the money
    utxos = utxos,               // unspent transaction outputs
    feeRate = feeRate,           // fee rate per byte
    changeSPK = creditingSPK,    // where to send the change
    network = networkParams      // bitcoin network information
  )

@@

  val tx: Transaction = Await.result(signedTxF, 1 second)
\end{lstlisting}

Line one to seven creates a transaction builder which is then signed
on line ten.  We can now use our transaction object on line twelve.
For example, after calling \emph{hex} on it, we can send the returned
string to the bitcoin network.

@@

Bitcoin-S offers a function called \emph{checkTransaction} located in the ScriptInterpreter object.
This is its type signature:
\begin{lstlisting}[style=scala]
  checkTransaction(transaction: Transaction): Boolean
\end{lstlisting}

\begin{lstlisting}[style=scala]
/**
  * Checks the validity of a transaction in accordance to bitcoin core's CheckTransaction function
  * https://github.com/bitcoin/bitcoin/blob/f7a21dae5dbf71d5bc00485215e84e6f2b309d0a/src/main.cpp#L939.
  */
def checkTransaction(transaction: Transaction): Boolean = {
  val inputOutputsNotZero =
    !(transaction.inputs.isEmpty || transaction.outputs.isEmpty)
  val txNotLargerThanBlock = transaction.bytes.size < Consensus.maxBlockSize
  val outputsSpendValidAmountsOfMoney = !transaction.outputs.exists(o =>
    o.value < CurrencyUnits.zero || o.value > Consensus.maxMoney)

  val outputValues = transaction.outputs.map(_.value)
  val totalSpentByOutputs: CurrencyUnit =
    outputValues.fold(CurrencyUnits.zero)(_ + _)
  val allOutputsValidMoneyRange = validMoneyRange(totalSpentByOutputs)
  val prevOutputs = transaction.inputs.map(_.previousOutput)
  val noDuplicateInputs = prevOutputs.distinct.size == prevOutputs.size

@@

@@

\todo{Kai: about errors, is it enough to reference to the chapter 4.2 where they are described??}
\todo{After integrating the Stainless plugin in the Bitcoin-S sbt project, 
there were many errors because of the different sbt versions.  
Some errors are described in the section 4.2.}
It takes too much time to fix them all so it should be easier to extract
the classes needed for the \texttt{checkTransaction} function.

@@

\todo{
kai text

extracted 2 files (they are in code/addition/src/main/scala/addition/original)

changes from original to reduced (new folder):
  - package name

  - removed numbers other than Int64

@@

\todo{ramon: write this section}
\todo{ramon: keep close to original code}
\todo{ramon: for every transformation: before and after}
\todo{ramon,anna: for every transformation: why does it preserve the semantics}
\todo{ramon,anna: add comments}
\todo{ramon,anna: explain what the code does and why}

@@

>>>>

%\usepackage{newtxtext}
%\usepackage{newtxmath}
\usepackage[zerostyle=b,scaled=.88]{newtxtt}

@@

\paragraph{Outline and Properties to Verify.} In the next section we
try to verify the property that a regular (non-coinbase) transaction
cannot generate new coins. We call it the \emph{no-inflation
  property}. Trying to verify it, we uncover and fix a bug in the
bitcoin-s library. We then find that there is too much code involved
that lies outside of the supported fragment to currently make this
verification feasible. Instead, we turn to a simpler property to
verify. The simplest possible property we can think of is the fact
that adding zero satoshis to a given amount of satoshis yields the
given amount of satoshis. We call it the \emph{addition-with-zero
  property} and we try to verify it in Section~3. Here as well we see
that a significant part of the code lies outside of the supported
fragment. However, after a series of equivalent transformations we
arrive at code that we successfully verify.

@@

A crucial function for the verification of the no-inflation property
is the \texttt{checkTransaction} function, shown in
Figure~\ref{fig:checktrans}. To better understand it, we first see how
to create a transaction.

\paragraph{Creating a Transaction}

% The bitcoin-s library has a bitcoin
% transaction builder class with the following constructor:

% \begin{lstlisting}[style=scala]
% def apply(
%       destinations: Seq[TransactionOutput],
%       utxos: BitcoinTxBuilder.UTXOMap,
%       feeRate: FeeUnit,
%       changeSPK: ScriptPubKey,
%       network: BitcoinNetwork): Future[BitcoinTxBuilder] = {...}
% \end{lstlisting}

% The purpose of the return type Future is not clear to us here, since
% the constructor always returns a completed future. This might be for
% future purposes.

To create a transaction, we first need some coins -- an unspent
transaction output. We could load an actual unspent transaction output
from the bitcoin network, but we create one manually in order to see
this process. So we first create an (invalid) transaction with one
output in Figure~\ref{fig:prevtx}.

\begin{figure}
\begin{lstlisting}[style=scala]
  val privKey = ECPrivateKey.freshPrivateKey
  val creditingSPK = P2PKHScriptPubKey(pubKey = privKey.publicKey)

@@

  \caption{Creating a transaction output to spend}
  \label{fig:prevtx}
\end{figure}

We first create a keypair, then a lock script with its public key,
then the amount of satoshis, then a transaction output (utxo) for that
amount and locked with that script. Finally we create the actual
transaction with that output and no inputs. Of course, that is not a
valid transaction, because it creates coins out of nothing. In
particular, checkTransaction(prevTx) is false.

Now that we have a transaction output, we create a new transaction to
spend it.

First, we need some out points.  They point to outputs of previous
transactions.  We use the index zero, because the previous transaction
has only one output that becomes the first index zero.  If there were
two previous outputs, the second output would become the index 1 and
so on.
\begin{figure}
\begin{lstlisting}[style=scala]
  val outPoint = TransactionOutPoint(prevTx.txId, UInt32.zero)

@@

  val utxos = List(utxoSpendingInfo)

  val destinationAmount = Satoshis(Int64(5000))

@@

  val destinations = List(
    TransactionOutput(currencyUnit = destinationAmount, scriptPubKey = destinationSPK)
  )

  val feeRate = SatoshisPerByte(Satoshis.one)

  val networkParams = RegTest // some static values for testing

  val txBuilderF: Future[BitcoinTxBuilder] = BitcoinTxBuilder(
    destinations = destinations, // where to send the money
    utxos = utxos,               // unspent transaction outputs
    feeRate = feeRate,           // fee rate per byte
    changeSPK = creditingSPK,    // where to send the change
    network = networkParams      // bitcoin network information
  )

@@

  val tx: Transaction = Await.result(signedTxF, 1 second)
\end{lstlisting}
  \caption{Creating a transaction}
  \label{fig:tx}
\end{figure}

This utxos are the inputs of our transaction.
Second, we need destinations to spend the bitcoins to.
For the sake of convenience we create only one.
We spend 5000 satoshis to the newly created random public key.
Finally, we define the fee rate in satoshis per one byte transaction
size as well as some bitcoin network parameters.  The bitcoin network
parameters are not important, so we use some static values normally
used when testing.

Now lets build the transaction with those data.
Line one to seven creates a transaction builder which is then signed
on line ten.  We can now use our transaction object on line twelve.
For example, after calling \emph{hex} on it, we can send the returned
string to the bitcoin network.

@@

Bitcoin-S offers a function called \emph{checkTransaction} located in the ScriptInterpreter object.

\begin{figure}
\begin{lstlisting}[style=scala]
/**
  * Checks the validity of a transaction in accordance to bitcoin core's CheckTransaction function
  * https://github.com/bitcoin/bitcoin/blob/f7a21dae5dbf71d5bc00485215e84e6f2b309d0a/src/main.cpp#L939.
  */
def checkTransaction(transaction: Transaction): Boolean = {
  val inputOutputsNotZero =
    !(transaction.inputs.isEmpty || transaction.outputs.isEmpty)
  val txNotLargerThanBlock = transaction.bytes.size < Consensus.maxBlockSize
  val outputsSpendValidAmountsOfMoney = !transaction.outputs.exists(o =>
    o.value < CurrencyUnits.zero || o.value > Consensus.maxMoney)

  val outputValues = transaction.outputs.map(_.value)
  val totalSpentByOutputs: CurrencyUnit =
    outputValues.fold(CurrencyUnits.zero)(_ + _)
  val allOutputsValidMoneyRange = validMoneyRange(totalSpentByOutputs)
  val prevOutputTxIds = transaction.inputs.map(_.previousOutput.txId)
  val noDuplicateInputs = prevOutputTxIds.distinct.size == prevOutputTxIds.size

@@

  \caption{The \texttt{checkTransaction} function in the \texttt{ScriptInterpreter} object}
  \label{fig:checktrans}
\end{figure}

@@

After integrating the Stainless plugin in the Bitcoin-S sbt project, 
there were many errors because of the different sbt versions.  
Some errors are described in the section 4.2.
It takes too much time to fix them all so it should be easier to extract
the classes needed for the \texttt{checkTransaction} function.

@@

The relevant code is in two files: \texttt{CurrencyUnits.scala} and \texttt{NumberType.scala}.

reference: code/addition/src/main/scala/addition/original

From those files we removed all code that is not needed for our
verification. For example, we removed all number types except for
Int64 (so Int32, UInt64, etc.) because we do not use them. We also
removed the superclasses Factory and NetworkElement of CurrencyUnit
and Number, respectively because the inherited members are not used by
the relevant code. Also, we removed all binary operations on Number that are not used.

@@

@@

\todo{conclusions: what's future work? how to
  change bitcoin-s? how to extend stainless?}
